External priors for the next generation of CMB experiments.

### 2015-11-18 Notes/N_eff_prior_paper_v2.tex
add omch2 right scaling in CMB, specified Planck prior (the same of what Dunkley used) and comment on Scott's comment

%\newcommand\sd[1]{{\bf SD: #1}}
\newcommand{\sd}[1]{\textcolor{red}{[{\bf SD}: #1]}}

@@

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe. They will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution expected in next generation, dubbed Stage IV (S4)~\cite{2013arXiv1309.5383A}.
The recent past is  reassuring. Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.

@@

For CMB experiments the Fisher matrix can be related to the power spectrum $\boldsymbol{C}_\ell$ by:
\begin{equation}
 F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{\rm sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right)
 \label{eqn:Fij_def2}
 \end{equation}
 where $f_{\rm sky}$, the fraction of sky covered, is set to 0.75 throughout.
In this work we constrain cosmological parameters with CMB temperature and E mode polarization together with the reconstructed lensing potential of large scale structure. Therefore, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
 \begin{eqnarray}
 	\centering
		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{T\phi} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{T\phi} & 0 & C_\ell^{\phi} + N_\ell^{\phi}\end{array}\right).
	\label{eqn:cov_definition}
\end{eqnarray}
The terms $N_\ell^{X}$ represent the instrumental noise power of the specific experiment and will be described at the end of this section.
Note that we are neglecting the term $C_\ell^{E\phi}$ since it  contains very little information while adding possible numerical issues~\cite{wu:2014,2013PhRvD..87h3008H}.
Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. \sd{Can we say something more positive about these effects? Why are we allowing ourselves to make this approximation?} Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.

@@

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{
Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. %($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ). 
Each prior is expressed in units of the error that would be obtained internally from the CMB experiment; e.g., $\sigma(H_0)_{\rm CMB}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%. $H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-like prior\sd{What does Planck prior mean here?}. The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
\label{fig:prior_omeganuh2}
\end{center}
\end{figure}
%==========

@@

\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment\sd{What does Planck prior mean?} %($N_{det}=10^{6}$, $f_{sky}=0.75$,  $2<\ell<5000$) 
as a function of external priors. Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}. 
There, the prior was described as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI. Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed are constrained by these low-redshift distances.
The improvement given by a prior on $\Omega_ch^2$ is also a consequence of the fact that the CMB lensing is sensitive to $f_\nu$; therefore in order to constrain $\Omega_{\nu}h^2$ from this ratio we need to determine the matter energy density. 
A prior on  $H_{0}$ helps with this as well since the effects of external priors are not completely independent.
The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^3$\sd{get the right power from the Planck paper and cite that} and for this reason too an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors.

@@

\begin{acknowledgments}
AM wants to thank Zhen Pan who allowed a careful cross-check of our results.
This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.
%%%=================================================================
The work of SD is supported by the U.S. Department of Energy, including grant DE-FG02-95ER40896.
\end{acknowledgments}

>>>>

%\newcommand\sd[1]{{\bf SD: #1}}
\newcommand{\sd}[1]{\textcolor{red}{[{\bf SD}: #1]}}
\newcommand{\am}[1]{\textcolor{blue}{[{\bf AM}: #1]}}

@@

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe. They will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution expected in next generation, dubbed Stage IV (S4)~\cite{2013arXiv1309.5383A}.
The recent past is  reassuring. Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} \am{it is ok for me to skip the list entirely otherwise feel free to add names I can add references later} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.

@@

For CMB experiments the Fisher matrix can be related to the power spectrum $\boldsymbol{C}_\ell$ by:
\begin{equation}
 F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{\rm sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right)
 \label{eqn:Fij_def2}
 \end{equation}
 where $f_{\rm sky}$, the fraction of sky covered, is set to 0.75 throughout.
In this work we constrain cosmological parameters with CMB temperature and E mode polarization together with the reconstructed lensing potential of large scale structure. Therefore, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
 \begin{eqnarray}
 	\centering
		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{T\phi} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{T\phi} & 0 & C_\ell^{\phi} + N_\ell^{\phi}\end{array}\right).
	\label{eqn:cov_definition}
\end{eqnarray}
The terms $N_\ell^{X}$ represent the instrumental noise power of the specific experiment and will be described at the end of this section.
Note that we are neglecting the term $C_\ell^{E\phi}$ since it  contains very little information while adding possible numerical issues~\cite{wu:2014,2013PhRvD..87h3008H}.
Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. 
\am{My understanding is that people do that: unlensed to avoid double counting in the informations with $C^{\phi}$, Aurelien and Wayne showed that unlensed + gaussian is similar to lensed + non-gaussian how big is the effect i do not know. Aurelien is writing a paper about that}
\sd{Can we say something more positive about these effects? Why are we allowing ourselves to make this approximation?} Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.

@@

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{
Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. %($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ). 
Each prior is expressed in units of the error that would be obtained internally from the CMB experiment; e.g., $\sigma(H_0)_{\rm CMB}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%. $H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-pol prior . The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
\label{fig:prior_omeganuh2}
\end{center}
\end{figure}
%==========

@@

\reffig{prior_omeganuh2} shows the errors on $\sum m_\nu$ with a CMB S4 + Planck experiment as a function of external priors. 
Our Planck prior corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}. 
There, the prior was described as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI. Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$ and $H_{0}$, which indeed are constrained by these low-redshift distances.
The improvement given by a prior on $\Omega_ch^2$ is also a consequence of the fact that the CMB lensing is sensitive to $f_\nu$; therefore in order to constrain $\Omega_{\nu}h^2$ from this ratio we need to determine the matter energy density. 
A prior on  $H_{0}$ helps with this as well since the effects of external priors are not completely independent.
The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} and for this reason too an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$.
As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors.

@@

\begin{acknowledgments}
We thank Wayne Hu for useful discussions. AM wants to thank Zhen Pan who allowed a careful cross-check of our results.
This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.
%%%=================================================================
The work of SD is supported by the U.S. Department of Energy, including grant DE-FG02-95ER40896.
\end{acknowledgments}
